Creating More Meaningful Explanations for Audio Deepfake Detection

### 2025-01-18 draft.tex
additions to the introduction and to the related work sections

	or predictions. This means that XAI systems can articulate why they made a particular
	decision, what factors influenced it, and how confident they are in their conclusion
	\cite{hind_explaining_2019}. For the detection of audio deepfakes, this could mean that the
	model can justify its classification with feature-based evidence, allowing for verification of
	the result by the end user. Much of the existing research, further discussed in the next
	section, follows the path of producing a model with the best possible benchmarks against
	datasets of samples, without taking into account if the results can be usefully interpreted,
	while existing explorations into making audio deepfake detection explainable have produced
	results that often require a lot of background knowledge to meaningfully interpret With the
	context of the research discussed in the next section, I would like to use new combinations of
	features with a novel model architecture to generate explanations for fake or real audio
	classification that are potentially more useful to their end user.

@@

@@

					li_comparative_2022} but slightly modified from the original concept, the HNRs
					are the ratios of the strengths of harmonic frequencies to the strengths of
					"noise", the total strength outside the harmonic frequencies. Unlike some
					previously seen examples \cite{chaiwongyen_contribution_2022,
					chaiwongyen_deepfake-speech_2023}, I calculate this ratio for each fundamental
					frequency length, instead of using one value for the whole sample. Given that
					 \(\gamma_{i}\) is the harmonic energy in a given fundamental frequency cycle
					 and \(\iota_{i}\) is the residual energy in a given fundamental frequency
					 cycle, the HNR at that cycle is calculated as follows:

>>>>

	or predictions \cite{hind_explaining_2019}. This means that XAI systems can articulate why
	they made a particular decision, what factors influenced it, and how confident they are in
	their conclusion \cite{hind_explaining_2019}. For the detection of audio deepfakes, this could
	mean that the model can justify its classification with feature-based evidence, allowing for
	verification of the result by the end user. Much of the existing research, further discussed
	in the next section, follows the path of producing a model with the best possible benchmarks
	against datasets of samples, without taking into account if the results can be usefully
	interpreted, while existing explorations into making audio deepfake detection explainable have
	produced results that often require a lot of background knowledge to meaningfully interpret.
	Aside from this, explainability in the area of audio deepfake detection remains an open
	challenge \cite{cuccovillo_open_2022}, one even promoted by regularatory bodies such as the 
	European Union with the "right to explain" \cite{goodman_european_2017}. With the context of
	the research discussed in the next section, I would like to use new combinations of features
	with a novel model architecture to generate explanations for fake or real audio classification
	that are potentially more useful to their end user.

@@

	One previous example of an implementation of an explainable model for use in audio deepfake
	detection came from Ge et al. \cite{ge_explaining_2024}, who explain feature influence on
	models using the SHAP (SHapley Additive exPlanations) method. They apply this method using
	log-scaled power spectrograms as a feature.
	Another example is presented by Haq et al. \cite{haq_multimodal_2023}, who use the changes in
	emotional state as an input feature and represent "unlikely" changes on a graph so that it can
	be understood by an eventual end user.

@@

					li_comparative_2022} but done in a sliding window fashion instead of on the
					whole file, the HNRs are the ratios of the strengths of harmonic frequencies
					to the strengths of "noise", the total strength outside the harmonic
					frequencies. Unlike some previously seen examples
					\cite{chaiwongyen_contribution_2022, chaiwongyen_deepfake-speech_2023}, I
					calculate this ratio for each fundamental frequency length, instead of using
					one value for the whole sample. Given that \(\gamma_{i}\) is the harmonic
					energy in a given fundamental frequency cycle and \(\iota_{i}\) is the
					residual energy in a given fundamental frequency cycle, the HNR at that cycle
					is calculated as follows:
